Galaxy image statistical generator: how to be confident?

### 2024-09-25 paper/main.tex
update denoiser performance figs. Add the GAN cosine simil. figure

The generative models assume that the observations $\bm{x}$ are described by a probability distribution $p(\bm{x})$ and aim to give a representation of such distribution.  It is also assumed that the data are i.i.d samples of $p(\bm{x})$.  In the following sections, the intention is more to give some background knowledge rather than describe in great details the different  architecture types and their implementations.

@@

Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}$  (aka \textit{latent variable}) whose distribution is chosen \textit{a priori} among a parametrized distribution family $\pi_{\bm{\theta}_1}(\bm{z})$ differentiable both with respect to $\bm{z}$ and $\bm{\theta}_1$. In general the latent space has a much lower dimensionality than the data space which imply a information compression. Then, it yields $p(\bm{x}) = p(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z})$. However, the likelihood  $p(\bm{x}|\bm{z})$ is also unknown which leads to the introduction of a parametrized version $p_{\bm{\theta_2}}(\bm{x}|\bm{z})$ with the differentiability properties as $\pi_{\bm{\theta}_1}$. Using the notation $\bm{\theta}=(\bm{\theta_1},\bm{\theta_2})$,  these parameters are adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $N$ data samples ($\{\bm{x}^{i}\}_{i<N}$). The best solution $\bm{\theta}_{ML}$ is the maximum of $p_{\bm{\theta}}(x)$ but in practice this leads to  intractable problem (either analytical or numerical) as one should evaluate the gradient with respect to $\bm{\theta}$ of the integral 

@@

It is usual to call $p_{\bm{\theta}_2}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part. The different VAE models differ by the distribution families used. The simplest choice may be commonly-used Gaussian multivariate distributions, and the centred isotropic with unit variance version ($\mathcal{N}(\bm{z},\bm{0},\bm{1}$) for the $\bm{z}$ prior (nb. $\bm{\theta}_1$ are missing is this scenario). Once the optimization is done, the \textit{decoder} is used as a $\bm{x}$ generator by feeding with $\bm{z}$ samples. It turns out that simple VAE tends to generate blurry images and may  suffer from the problem of \textit{posterior collapse} which motivate active ML researches \citep[e.g.][]{engel2018latent,Takida2022} and modified VAE versions to overcome this issue as for instance in \cite{Lanusse2021}. 
%
\subsection{Generative Adversarial Networks}
%
The vanilla GAN \citep{goodfellow2014generative} shares with VAE the target to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}$ \textit{prior} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$. To do so, a second network ($\bm{D}$) acts as a \textit{discriminator} which aims to state if a sample $\bm{x}$ is from the model distribution or the data distribution. To reach that goal one solves the \texttt{min-max} problem
\begin{equation}
\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\}
\end{equation} 
where $p_{data}$ is the data generating distribution. Both the  $\bm{G}$ and $\bm{D}$ are parametrized networks. Notice that at parameter of $\bm{G}$ fixed, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression
\begin{equation}
\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})}
\end{equation}
with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}} |^{-1}$ using the Jacobian of the \textit{generator}, leading to a reformulation of the \texttt{min-max} problem as followed
\begin{equation}
\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right) }  + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\}
\end{equation}
leading to the optimal solution $p_{\bm{G}} = p_{data}$ yielding a perfect matching of the data distribution.  But, training GAN model via the above objective function is in general unstable with a pathology called \textit{mode-collapse} in the literature, dealing with the fact that the \textit{discriminator} may overfit the dataset too quickly leading to vanishing gradients \citep{Gulrajani2017}. This training instability has been studied by many authors \citep[see e.g.][for some reviews]{Saxena2021,Jozdani2022}. Architectures like \texttt{BigGAN} \citep{Brock2019} and \texttt{StyleGAN2} \cite{Karras2018ASG,Karras2020} overcome the gradient flow issue by different techniques beeing "state-of-the-art" improvements but at the expense of computational resources.

@@

%Coccomini2021 utilise Lightweight GAN
% Schawinski2017,Hemmati_2022 GAN vanilla
GAN architectures have been used for galaxy generation in conjunction of deblending and deconvolving tasks: for instance \cite{Schawinski2017,Hemmati_2022} use vanilla-GANs while  \cite{Coccomini2021} the \texttt{light-weight-GAN}.
%
\subsection{Normalizing Flows}
%
Looking at Equation \ref{eq-VAE-ELBO}, one notices that the optimal solution is reached in case of $\DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))=0$ which is true iif $p_{\bm{\phi}}(\bm{z}|\bm{x}) = p(\bm{z}|\bm{x})$. But this situation cannot be true in general as typical $p_{\bm{\phi}}$ parametrizations involve for instance independent Gaussian distributions. The limited choices of the parametrization is one of the limitation of VAE. So, one motivates the use of normalizing flows (NF) to allow for more flexibility in the distribution family design to hope that the true posterior is member of such family \citep{Tabak2010, Tabak2013a, Rezende2015}.

@@

The structure of the transformation (e.g. Eq.~\ref{eq-NVP_forward}) is a single split of the $\bm{z}$ vector coordinates in two parts: the first part $(z_{\leq s})$ is not transformed while the transformation of the second part ${z_{>s}})$ uses a function depending only on the first part. Notice that the conditioner functions do depends only on $(z_{\leq s})$ for both the forward or the backward transformations. This single split schema is illustrated on Figure \ref{fig-nvp}. 
%
\begin{figure}
\centering
\includegraphics[width=0.45\columnwidth]{fig-nvp.pdf}
\includegraphics[width=0.45\columnwidth]{fig-nvp-inverse.pdf}
\caption{Affine coupling layer schema of forward and backward directions (Eqs.~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \cite{Papamakarios2021}).}
\label{fig-nvp}
\end{figure}
%
This splitting schema implemented in a so-called \textit{coupling layer} makes the flow-based model computationally efficient and using deep neural networks, one can achieve complex $(\bm{\alpha},\bm{\beta})$ functions. Now, designers of complete architectures can compose different \textit{coupling layers} by reversing $\bm{z}$ elements between layers to ensure that all elements are transformed at the end of the set of operations. An generalisation of any permutation ordering of the inputs can be performed using a $1\times 1$ convolution layer. \cite{2016arXiv160508803D} improves the modelling using a multi-scale approach which would be prohibitive to describe here, and \cite{Kingma2018} use all these extensions to build the \texttt{Glow} model whose schema is displayed on Figure \ref{fig-Glow-archi}.
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[b]{0.45\textwidth}
    \begin{center}
        \centering
        \includegraphics[width=.7\textwidth]{fig-Glow-1.pdf}
        \caption{One step of Glow flow.}
    \end{center}
    \end{subfigure}%
    \vspace{5mm}
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=.99\textwidth]{fig-Glow-2.pdf}
        \caption{Multi-scale architecture \citep{2016arXiv160508803D}.}
    \end{subfigure}
    \caption{Figure from \citep{Kingma2018}. The flow step is composed of a scale and bias layer with data dependent initialization (\texttt{actnorm}), a $1\times 1$ invertible convolution layer and an affine coupling layer with $(\bm{\alpha},\bm{\beta})$ issued from a shallow convolutional neural network. Both $\bm{x}$ and $\bm{y}$ are tensors of
shape $[h \times w \times c]$ with spatial dimensions $(h, w)$ and channel dimension $c$ ans $(i, j)$ denote
spatial indices into tensors $\bm{x}$ and $\bm{y}$.  This flow step is embedded in a multi-scale  architecture which has $K$ flow steps and a number of levels $L$ \citep{2016arXiv160508803D}. The squeezing operation reduces a $s \times s \times c$ tensor  into a $s/2\times s/2 \times 4c$ tensor which mix spatial and channel components. At each scale, each channel image is divided into $2\times 2$ patches ($s=2$) and each patch is squeezed. }
    \label{fig-Glow-archi}
\end{figure}
 %
\subsection{Score based diffusion models}
%
To generate samples from $p(\bm{x})$, we can proceed as follows \citep[e.g][]{Chang2023,LinYang2023}. Starting from a sample $\bm{x}_0$ assumed to be drawn from the distribution $p_0(\bm{x})=p(\bm{x})$, we progressively transform it until the  distribution is as simple as for instance $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.  The idea is then to reverse the process (the \textit{backward} mode) by generating a new sample from a new instance of noise. This transfer mapping looks very similar in concept as normalizing flows. Their latent spaces have the same dimension as the data space, there is no compression as for VAE architecture, but the main difference consists in the fact that the normalizing flows are deterministic by essence while the \textit{score diffusion} models are stochastic.

@@

Denoising diffusion stochastic models have gain in popularity due to the fact that they get as good as results as GAN and have certain advantages. Notably they do not need any adversarial networks, the mathematical ground is clear and robust, the training is also theoretically well defined. But they use an iterative schema for the generation that can be computationally intensive and slower compared to GAN, and the implementation of stochastic processes may require some theoretical expertise even if this argument can be mitigated. More discussion about the comparison between the two architectures  can be found in this article \cite{dhariwal2021diffusion}. We do not investigate further the mixing of flow-based and diffusion-based models as \citep[e.g.][]{zhang2021diffusion,gong2021interpreting} as well as the use of diffusion models to improve GAN stability as introduced by \cite{Wang2022}. This can be envisaged in a future study.
%
\section{Experiment}
%
Having briefly outlined different architectures of generative models in the previous section it turns out that one aspect in common between diffusion-based and flow-based models consists in the generation process of a $\bm{x}$ sample which starts by sampling a centred isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{z},\bm{0},\bm{1})$ (i.e. a white noise) of dimension $d$ equal to the final sample. But a question rises: if we train two generative models of the same kind, let say two flow-based or two diffusion-based, and if we feed them with the same white noise image, does the generative process of the two models give the same final image? If not we can ask ourselves about the nature of the learned process and if we are faced to a real probability density $p(\bm{x})$ sample or does the result is a sort of mixture of the original dataset images?

An answer to the previous question has been given in \cite{kadkhodaie2024generalization} and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes. We propose to use the same methodology by comparing typically two generative models of same kind trained with two different non-overlapping sub-datasets extracted from the same dataset of original images.

%
\subsection{Datasets}
%
Concerning the training dataset, \cite{kadkhodaie2024generalization}  have used different sources commonly-used in ML as \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets. All dataset images use  gray level single channel ranging in $[0,1]$.  The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.

We are more concerned with galaxy image generation, but as we need to train two models with up to at least $100,000$ images each, we have chosen to use the SDSS (Sloan Digital Sky Survey) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} with some modification on the preprocessing. First, all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels, and secondly from the $(z,g,r)$ channels we produce a single  channel in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}. Moreover, the gray levels are coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}. With such dataset according to \cite{kadkhodaie2024generalization} results, we expect the transition between memorization to generalisation regimes to occur with a number of training images in the range $10^4-10^5$.

\subsection{Models}
%
{\color{red} Concerning the models, for the GAN model type ....  
For the diffusion-based model, we have used the same \texttt{U-Net} architecture \citep{2015arXiv150504597R}  as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{Source code:  \url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. Second, for comparison purposes, we use the \texttt{Glow} flow-based model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}  \citep{PyTorch2019}. The architecture has been presented on Figure \ref{fig-Glow-archi} and we use $K=32$ and $L=4$.}

So, we have trained two \texttt{U-Net}-based diffusion models with independent datasets of sizes  $N=10^2,10^3,10^4,10^5$ extracted from the whole dataset composed of $250,000$ galaxy images. The same training procedure used by \cite{kadkhodaie2024generalization} has been applied. Concerning the \texttt{Glow} model, we use two independent datasets of sizes $N=10^5$ and train two models with  batches of size $32$ for $450,000$ iterations. {\color{red} We refer the reader to the  github repository companion of the present article for the details of the different codes and notebooks used and models trained as well as to get the whole dataset of $250,000$ original galaxy images.}
\subsection{Results}
\subsubsection{Generated images}
Figure \ref{fig-Original-Glow-UNet-Gan-samples} presents some galaxy images extracted from the original dataset and generated by  diffusion-based and flow-based models trained with $N=10^5$ original images. 
\begin{figure}
    \centering
        \includegraphics[width=0.8\textwidth]{fig-model_sample_images.pdf}
    \caption{Some \textit{validation} images from the original dataset (first row from top) and  generates samples from different models trained with $N=10^5$ images:  the \texttt{light-weight} GAN-based model (second row), the \texttt{Glow} flow-based model (third row) and the diffusion-based with \texttt{U-Net} denoiser (bottow row).}
    \label{fig-Original-Glow-UNet-Gan-samples}
\end{figure}
At first look the generated images may be not distinguishable from the original images.
To be a bit more quantitative, Figure \ref{fig-morpho-coeff} presents  some morphological coefficients that have been computed using \texttt{Statmorph}\footnote{\url{https://github.com/vrodgom/statmorph}} \citep{2019MNRAS.483.4140R}: the Concentration, Asymmetry and Smoothness \citep{2000AJ....119.2645B,2003ApJS..147....1C,2004AJ....128..163L}  as well as the Gini and M20 coefficients \citep{2004AJ....128..163L,10.1093/mnras/stv2078}.
\begin{figure}
    \centering
		\includegraphics[width=0.5\textwidth]{fig-CAS-all.pdf}
		\includegraphics[width=0.35\textwidth]{fig-gini-m20-all.pdf}
	\caption{Corner plots\protect\footnotemark of morphological coefficients computed with $10^4$ images, first from the original dataset but \textit{never} used by any models (red, "test"), second from a diffusion-based model generation (blue, "diffusion") and from a flow-based model generation (green, "Glow") (both models used for Figure \ref{fig-Original-Glow-UNet-samples}). The figure has been inspired by \cite{HACKSTEIN2023100685}.}
	\label{fig-morpho-coeff}
\end{figure}
\footnotetext{Plots from the library \url{https://corner.readthedocs.io} \citep{corner2016}.}
The distributions derived from the \texttt{U-Net} diffusion model are in better agreement with those obtained with the original images than are those derived from the \texttt{Glow} flow model, although this is not dramatic.
Notice that we do not expect high quality images as well as state-of-the-art competitive results, see for instance \citep{ravanbakhsh2016,Fussell2019,Lanusse2021,smith2021,HACKSTEIN2023100685} for better image quality. Our goal is somewhat different: we ask ourselves if with such low resolution and low complexity images, does the models of the same kind learn the same thing with moderate dataset size?
\subsubsection{\texttt{U-Net} denoising performances}

>>>>

The generative models assume that the observations $\bm{x}\in \mathbb{R}^d$ (in the context of image modelling $d$ is the number of pixels  times the number of channels) are described by a probability distribution $p(\bm{x})$ and aim to give a representation of such distribution.  It is also assumed that the data are i.i.d samples of $p(\bm{x})$.  In the following sections, the intention is more to give some background knowledge rather than describe in great details the different  architecture types and their implementations.

@@

Using \texttt{VAE} \citep{Kingma2014} means that it is assumed that there exists an underlying unobservable stochastic variable $\bm{z}\in\mathbb{R}^{d_\ell}$  (aka \textit{latent variable}) whose distribution is chosen \textit{a priori} among a parametrized distribution family $\pi_{\bm{\theta}_1}(\bm{z})$ differentiable both with respect to $\bm{z}$ and $\bm{\theta}_1$. In general the latent space has a much lower dimensionality than the data space which imply a information compression ($d_\ell \ll d$). Then, it yields $p(\bm{x}) = p(\bm{x}|\bm{z})\pi_{\bm{\theta}_1}(\bm{z})$. However, the likelihood  $p(\bm{x}|\bm{z})$ is also unknown which leads to the introduction of a parametrized version $p_{\bm{\theta_2}}(\bm{x}|\bm{z})$ with the differentiability properties as $\pi_{\bm{\theta}_1}$. Using the notation $\bm{\theta}=(\bm{\theta_1},\bm{\theta_2})$,  these parameters are adjusted in such a way that  $p_{\bm{\theta}}(x)$ matches the empirical distribution of the $N$ data samples ($\{\bm{x}^{i}\}_{i<N}$). The best solution $\bm{\theta}_{ML}$ is the maximum of $p_{\bm{\theta}}(x)$ but in practice this leads to  intractable problem (either analytical or numerical) as one should evaluate the gradient with respect to $\bm{\theta}$ of the integral 

@@

It is usual to call $p_{\bm{\theta}_2}(\bm{x}|\bm{z})$ the \textit{decoder} part of the whole model while $p_{\bm{\phi}}(\bm{z}|\bm{x})$ is the \textit{encoder} part. The different VAE models differ by the distribution families used. The simplest choice may be commonly-used Gaussian multivariate distributions, and the centred isotropic with unit variance version ($\mathcal{N}(\bm{z},\bm{0},\bm{1}$) for the $\bm{z}$ prior (nb. $\bm{\theta}_1$ are missing is this scenario). Once the optimization is done, the \textit{decoder} is used as a $\bm{x}$ generator by feeding with $\bm{z}$ samples. It turns out that simple VAE tends to generate blurry images and may  suffer from the problem of \textit{posterior collapse} which motivate active ML researches \citep[e.g.][]{engel2018latent,Takida2022} and modified VAE versions to overcome this issue as for instance in \cite{Lanusse2021}. 
%
\subsection{Generative Adversarial Networks}
\label{sec-GAN}
%
The vanilla GAN \citep{goodfellow2014generative} shares with VAE the purpose to optimize a \textit{generator} network ($\bm{G}$) to get $\bm{x}$ samples from a latent variable $\bm{z}$ such that  $\bm{x} = \bm{G}(\bm{z})$ with $\bm{z}\in\mathbb{R}^{d_\ell}$ (in general $d_\ell\ll d$) \textit{prior} $\pi(\bm{z})$ is for instance a Gaussian distribution $\mathcal{N}(\bm{z},\bm{0},\bm{1})$. To do so, a second network ($\bm{D}$) acts as a \textit{discriminator} which aims to state if a sample $\bm{x}$ is from the model distribution or the data distribution. To reach that goal one solves the \texttt{min-max} problem
\begin{equation}
\min_{\bm{G}} \max_{\bm{D}}\left\{ \Esp_{\bm{x}\sim p_{data}(\bm{x})}[\log \bm{D}(\bm{x})] + \Esp_{\bm{z}\sim \pi(\bm{z})}[\log(1-\bm{D}(\bm{G}(\bm{z})))] \right\}
\end{equation} 
where $p_{data}$ is the data generating distribution. Both the  $\bm{G}$ and $\bm{D}$ are parametrized networks. Notice that at parameter of $\bm{G}$ fixed, the optimal \textit{discriminator} $\bm{D}^\ast$ has the following expression
\begin{equation}
\bm{D}^\ast(\bm{x}) = \frac{p_{data}(\bm{x})}{p_{data}(\bm{x}) + p_{\bm{G}}(\bm{x})}
\end{equation}
with $p_{\bm{G}}(\bm{x}) = \pi(\bm{z})|\det J_{\bm{G}} |^{-1}$ using the Jacobian of the \textit{generator}, leading to a reformulation of the \texttt{min-max} problem as followed
\begin{equation}
\min_G \left\{ \DKL{\left( p_{data} \left\| \frac{p_{data} + p_{\bm{G}}}{2}\right. \right) }  + \DKL{\left(p_{\bm{G}} \left\| \frac{p_{data} + p_{\bm{G}}}{2} \right. \right)} \right\}
\end{equation}
leading to the optimal solution $p_{\bm{G}} = p_{data}$ yielding a perfect matching of the data distribution.  But, training GAN model via the above objective function is in general unstable with a pathology called \textit{mode-collapse} in the literature, dealing with the fact that the \textit{discriminator} may overfit the dataset too quickly leading to vanishing gradients \citep{Gulrajani2017}. This training instability has been studied by many authors \citep[see e.g.][for some reviews]{Saxena2021,Jozdani2022}. Architectures like \texttt{BigGAN} \citep{Brock2019} and \texttt{StyleGAN2} \cite{Karras2018ASG,Karras2020} overcome the gradient flow issue by different techniques beeing "state-of-the-art" improvements but at the expense of computational resources.

@@

%Coccomini2021 utilise Lightweight GAN
% Schawinski2017,Hemmati_2022 GAN vanilla
GAN architectures have been used for galaxy generation in conjunction of deblending and deconvolving tasks: for instance \cite{Schawinski2017,Hemmati_2022} use vanilla-GANs while  \cite{Coccomini2021} the \texttt{light-weight-GAN}.
%
\subsection{Normalizing Flows}
\label{sec-NF}
%
Looking at Equation \ref{eq-VAE-ELBO}, one notices that the optimal solution is reached in case of $\DKL(p_{\bm{\phi}}(\bm{z}|\bm{x})\| p(\bm{z}|\bm{x}))=0$ which is true iif $p_{\bm{\phi}}(\bm{z}|\bm{x}) = p(\bm{z}|\bm{x})$. But this situation cannot be true in general as typical $p_{\bm{\phi}}$ parametrizations involve for instance independent Gaussian distributions. The limited choices of the parametrization is one of the limitation of VAE. So, one motivates the use of normalizing flows (NF) to allow for more flexibility in the distribution family design to hope that the true posterior is member of such family \citep{Tabak2010, Tabak2013a, Rezende2015}.

@@

The structure of the transformation (e.g. Eq.~\ref{eq-NVP_forward}) is a single split of the $\bm{z}$ vector coordinates in two parts: the first part $(z_{\leq s})$ is not transformed while the transformation of the second part ${z_{>s}})$ uses a function depending only on the first part. Notice that the conditioner functions do depends only on $(z_{\leq s})$ for both the forward or the backward transformations. This single split schema is illustrated on Figure \ref{fig-nvp}. 
%
\begin{figure}
\centering
\includegraphics[width=0.45\columnwidth]{fig-nvp.pdf}
\includegraphics[width=0.45\columnwidth]{fig-nvp-inverse.pdf}
\caption{Affine coupling layer schema of forward and backward directions (Eqs.~\ref{eq-NVP_forward},\ref{eq-NVP_backward}) with a single split mechanism (figure inspired by \cite{Papamakarios2021}).}
\label{fig-nvp}
\end{figure}
%
This splitting schema implemented in a so-called \textit{coupling layer} makes the flow-based model computationally efficient and using deep neural networks, one can achieve complex $(\bm{\alpha},\bm{\beta})$ functions. Now, designers of complete architectures can compose different \textit{coupling layers} by reversing $\bm{z}$ elements between layers to ensure that all elements are transformed at the end of the set of operations. An generalisation of any permutation ordering of the inputs can be performed using a $1\times 1$ convolution layer. \cite{2016arXiv160508803D} improves the modelling using a multi-scale approach which would be prohibitive to describe here, and \cite{Kingma2018} use all these extensions to build the \texttt{Glow} model whose schema is displayed on Figure \ref{fig-Glow-archi}.
\begin{figure}
    \centering
    \vspace{-0cm}
    \begin{subfigure}[b]{0.45\textwidth}
    \begin{center}
        \centering
        \includegraphics[width=.7\textwidth]{fig-Glow-1.pdf}
        \caption{One step of Glow flow.}
    \end{center}
    \end{subfigure}%
    \vspace{5mm}
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=.99\textwidth]{fig-Glow-2.pdf}
        \caption{Multi-scale architecture \citep{2016arXiv160508803D}.}
    \end{subfigure}
    \caption{Figure from \citep{Kingma2018}. The flow step is composed of a scale and bias layer with data dependent initialization (\texttt{actnorm}), a $1\times 1$ invertible convolution layer and an affine coupling layer with $(\bm{\alpha},\bm{\beta})$ issued from a shallow convolutional neural network. Both $\bm{x}$ and $\bm{y}$ are tensors of
shape $[h \times w \times c]$ with spatial dimensions $(h, w)$ and channel dimension $c$ ans $(i, j)$ denote
spatial indices into tensors $\bm{x}$ and $\bm{y}$.  This flow step is embedded in a multi-scale  architecture which has $K$ flow steps and a number of levels $L$ \citep{2016arXiv160508803D}. The squeezing operation reduces a $s \times s \times c$ tensor  into a $s/2\times s/2 \times 4c$ tensor which mix spatial and channel components. At each scale, each channel image is divided into $2\times 2$ patches ($s=2$) and each patch is squeezed. }
    \label{fig-Glow-archi}
\end{figure}
 %
\subsection{Score based diffusion models}
\label{sec-diff}
%
To generate samples from $p(\bm{x})$, we can proceed as follows \citep[e.g][]{Chang2023,LinYang2023}. Starting from a sample $\bm{x}_0$ assumed to be drawn from the distribution $p_0(\bm{x})=p(\bm{x})$, we progressively transform it until the  distribution is as simple as for instance $\mathcal{N}(\bm{z},\bm{0},\bm{1})$.  The idea is then to reverse the process (the \textit{backward} mode) by generating a new sample from a new instance of noise. This transfer mapping looks very similar in concept as normalizing flows. Their latent spaces have the same dimension as the data space, there is no compression as for VAE and GAN architectures, but the main difference consists in the fact that the normalizing flows are deterministic by essence while the \textit{score diffusion} models are stochastic.

@@

Denoising diffusion stochastic models have gain in popularity due to the fact that they get as good as results as GAN and have certain advantages. Notably they do not need any adversarial networks, the mathematical ground is clear and robust, the training is also theoretically well defined. But they use an iterative schema for the generation that can be computationally intensive and slower compared to GAN, and the implementation of stochastic processes may require some theoretical expertise even if this argument can be mitigated as expertise is also needed for all kinds of architectures. More discussion about the comparison between the two architectures  can be found in this article \cite{dhariwal2021diffusion}. We do not investigate further the mixing of flow-based and diffusion-based models as \citep[e.g.][]{zhang2021diffusion,gong2021interpreting} as well as the use of diffusion models to improve GAN stability as introduced by \cite{Wang2022}. This can be envisaged in a future study.
%
\section{Experiment}
%
Having briefly outlined different architectures of generative models in the previous section, it turns out that the generation process of a $\bm{x}$ sample may start\footnote{nb. it is the typical implementation choice but one would have chosen uniform distribution as well.} by sampling a centred isotropic Gaussian distribution $\bm{z}\sim \mathcal{N}(\bm{z},\bm{0},\bm{1})$ (i.e. a white noise) of dimension $d_\ell$. In the case of VAE/GAN-based models $d_\ell\ll d$ while $d=d_\ell$ by construction for flow-based and diffusion-based models. But a question arises: if we train two generative models of the same kind, and if we feed them with the same white noise latent array, does the generative process of the two models give the same final image? If not we can ask ourselves about the nature of the learned process and if we are faced to a real probability density $p(\bm{x})$ sample or does the result is a sort of mixture of the original dataset images?

An answer to the previous question has been given in \cite{kadkhodaie2024generalization} in the context of diffusion-based models and point out the size of the dataset in the passage from \textit{memorization} to \textit{generation} regimes. We propose to use the same methodology by comparing typically two generative models of same kind trained with two different non-overlapping sub-datasets extracted from the same dataset of original images. In this task we cannot be exhaustive and choose only three models and a single dataset in our experiment. We think it is enough to draw some guide lines.
%
\subsection{Dataset}
%
Concerning the training dataset, \cite{kadkhodaie2024generalization}  have used different sources commonly-used in ML as \texttt{CelebA} \citep{Liu2015}, \texttt{LSUN} bedroom \citep{Yu2015} and  \texttt{CelebA HQ} \citep{KarrasALL18} datasets. All dataset images use  gray scale (single channel) ranging in $[0,1]$.  The \texttt{LSUN}  images have been down sampled to $80\times 80$ as well as the \texttt{CelebA} images which also have been further down sampled to $32\times 32$ resolution, while \texttt{CelebA HQ}  have been down sampled to $40\times 40$ resolution.

We are more concerned with galaxy image generation, but as it will be clear in the result section, we need to train two models with up to at least $100,000$ images each, so we have chosen to use the Sloan Digital Sky Survey (SDSS) dataset described in \citep{smith2021}\footnote{See \url{https://github.com/Smith42/astroddpm} to obtain the raw data.} with some modification on the preprocessing. First, all images are cropped about the target coordinate galaxy to a shape of $64\times 64$ pixels, and secondly from the $(z,g,r)$ channels we produce a single  channel in the range $[0,1]$ using \texttt{scipy} \citep{2020SciPy-NMeth} library\footnote{In particular the \texttt{make\_lupton\_rgb} function has been applied with $Q=8$ and $stretch=0.2$ setting}. %Moreover, the gray levels are coded using 5-bit precision in the line of experiment described in \cite{Kingma2018}.
 With such dataset according to \cite{kadkhodaie2024generalization} results, we expect the transition between memorization to generalisation regimes to occur with a number of training images in the range $10^4-10^5$.

\subsection{Models}
%
Concerning the models, we have chosen three architectures: a GAN, a flow-based and a diffusion-based.

For GAN model, we have chosen the \texttt{light-weight} architecture described in section \ref{sec-GAN} implemented\footnote{\url{https://github.com/lucidrains/lightweight-gan/}. We have chosen this code has it is more user-friendly compared to the original code given by authors (\url{https://github.com/odegeasslbc/FastGAN-pytorch/tree/main}).} in \texttt{PyTorch} \citep{PyTorch2019}. We have used default architecture settings as the latent space dimension ($d_\ell=256$) as well as the \textit{generator} and \textit{discriminator} (\textit{encoder} and \textit{decoders}) details. The total number of model parameters is about $36$~million.  We do not change as well the default optimizer choice and learning settings, except that first we force any input image to be resized to $128^2$, second we allow only horizontal and vertical flips for input data augmentation. Such model optimization takes about 30~hours for about $100,000$ iterations on a single Nvidia's GPU V100.

Concerning flow-based model, we use the \texttt{Glow} model implemented\footnote{Source code: \url{https://github.com/rosinality/glow-pytorch}} in \texttt{PyTorch}. The architecture has been presented on Figure \ref{fig-Glow-archi} (Sec.~\ref{sec-NF}) and we use $K=32$ and $L=4$ leading to about $61$~million parameters. Concerning the training we use defaults settings and as for the GAN training we allow only horizontal and vertical flips for input data augmentation. For completeness, the image pixel value encoding as been switched from 8-bit to 5-bit following \cite{Kingma2018}. We think that this has a minor impact on our results. We ask for $450,000$ iterations to complete a single model optimization which last approximately 100~hours.

For the diffusion-based model, we have used the same \texttt{U-Net} architecture\footnote{Notice that \cite{2015arXiv150504597R} have also studied a different \textit{denoiser} architecture leading to the same result.} \citep{2015arXiv150504597R} as \textit{denoiser} consisting of a contracting part and an expending part that allows for multi-scale architecture with scale correlations. Details are found in \cite{kadkhodaie2024generalization}\footnote{\url{https://github.com/LabForComputationalVision/memorization_generalization_in_diffusion_models}}. The number of parameters is about $7.6$~million.  The same training procedure used by \cite{kadkhodaie2024generalization} has been applied with 100 epochs, lasting approximately 100~hours.
